Analysing number of case and control based on new harmonised definition

### 2017-04-06 code/Rmd/harmonisedCaseControl.Rmd
Added markdown to harmonise case control definitions

---
title: "Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
                                 parentId = "syn7554308",
                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

@@

cl = makeCluster(detectCores()-2)
doParallel::registerDoParallel(cl)

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn7554308'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MAYO reprocessed RNASeq data';

thisFileName <- 'MAYO_REPROCESSED.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')

@@

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.

Following samples were removed `r indRemoved`

### Predict missing PMI
```{r predict.pmi}
PMI = METADATA[,c('PMI', 'ID'), drop = F]
PMI$PMI = PMI$PMI * 60
rownames(PMI) = METADATA$ID

# Get genes that are predictors of PMI and use them as control probes
CONTROL.PROBES_ID = 'syn6145639'
ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F) %>%
  dplyr::mutate(ID = x) %>%
  tidyr::separate(ID, c('ensembl_gene_id', 'position'), sep = '\\.')

# Conditional quantile normalisation
tmp.count = COUNT
genesToAnalyze = unique(intersect(rownames(GENE.GC)[!is.na(GENE.GC$percentage_gc_content)], 
                                  rownames(GENE.LEN)[!is.na(GENE.LEN$gene.length)]))
tmp.count = tmp.count[rownames(tmp.count) %in% genesToAnalyze,]
CQN.GENE_EXPRESSION = cqn(tmp.count,
                          x = GENE.GC[rownames(tmp.count), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(tmp.count), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Initial normalisation of gene expression
ind = rownames(CQN.GENE_EXPRESSION$y) %in% CONTROL.PROBES$ensembl_gene_id 
EXPR = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
EXPR = EXPR[ind,]
PMI = PMI[colnames(EXPR),,drop=F]

expr = dlply(PMI, .(ID), function(mtd, expr){
  if (dim(mtd)[1] > 1)
    expr = rowMeans(expr[, mtd$ID]) %>% rownameToFirstColumn('GeneID')
  else
    expr = expr[, mtd$ID] %>% rownameToFirstColumn('GeneID')
  expr = expr %>% plyr::rename(c('DF' = unique(mtd$ID)))
  return(expr)
}, EXPR) %>%
  join_all
rownames(expr) = expr$GeneID

PMI = PMI[,colnames(PMI) %in% c('ID','PMI')] %>% unique
rownames(PMI) = PMI$ID

PMI = PMI[intersect(rownames(PMI), colnames(expr)), ]
expr = expr[,intersect(rownames(PMI), colnames(expr))]

# Split data into trainining and prediction set
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]

plr.model = cv.glmnet(t(x.train), y.train$PMI, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
mse = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train$PMI)^2)/length(y.train$PMI)
y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('ID') %>%
  plyr::rename(c('1' = 'PMI'))

PMI = rbindlist(list(y.train, y.predict), use.names = T, fill = T)
PMI$PMI[PMI$PMI < 0] = 0

METADATA = METADATA %>% 
  dplyr::select(-PMI) %>%
  left_join(PMI)
```

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- "BrainRegion.Diagnosis"
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$ID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=6, fig.height=6}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

```{r data.explore, fig.width = 12, fig.height = 8}
my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# AgeAtDeath
p[[2]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(AgeAtDeath))) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PMI))) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + my.theme

# Ribosomal bases
p[[4]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme

# Intronic bases
p[[5]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme

# Intergenic bases
p[[6]] = ggplot(COVARIATES, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme

multiplot(plotlist = p, cols = 3)
```
### CPM Normalisation
Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene.length)]) %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region),
                Source = factor(Source))
plotdata$SampleID[!(plotdata$SampleID %in% c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX"))] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preadj.covariates.plot, fig.width=12, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix.

NOTE:
1. Using a mixed effect model where random effect is chosen as Donor_ID
2. Adding Source, FLOWCELL and Gender a priori to variable selection
3. Primary variable of interest BrainRegion.Diagnosis is excluded from the pool of available covariates for selection
```{r iterative.norm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Source', 'FLOWCELL', 'Gender');

# Assign residual covariates
residualCovars = setdiff(preAdjustedSigCovars$significantCovars, 
                         c(postAdjustCovars, primaryVariable, 'Donor_ID'))
residualSigCovars = preAdjustedSigCovars
covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and Donor_ID as random effect'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$Donor_ID)

  # Re-estimate voom weights
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                              design=DM1$design, plot=F, 
                              block = COVARIATES$Donor_ID, 
                              correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights,
                            block = COVARIATES$Donor_ID, 
                            correlation = correlation$cor)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates),
                         c(postAdjustCovars, primaryVariable, 'Donor_ID'))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```

### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2.1, fig.height=10, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region),
                Source = factor(Source))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```

```{r decompse.normalise.data2.2, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis (with BrainRegion.Diagnosis as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above

Interpretation of BrainRegion.Diagnosis
1. TCX.PSP: Progressive Supranuclear Palsy samples from Temporal cortex region
2. TCX.AD: Alzheimers Disease samples from Temporal cortex region
3. TCX.Control: Control samples from Temporal cortex region
4. TCX.Pathologic_Aging: Pathological Aging samples from Temporal cortex region
5. CER.PSP: Progressive Supranuclear Palsy samples from Cerebellum region
6. CER.AD: Alzheimers Disease samples from Cerebellum region
7. CER.Control: Control samples from Cerebellum region
8. CER.Pathologic_Aging: Pathological Aging samples from Cerebellum region

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp, fig.height=6, fig.width=18}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$Donor_ID, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
FIT = lmFit(VOOM.WEIGHTS)

>>>>

---
title: "Analysing number of case and control based on new harmonised definition"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

@@

@@

### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')

@@

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]

all.counts = list()
all.mtd = list()

all.counts$MAYO = COUNT
all.mtd$MAYO = METADATA
```

### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn7994853';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)

# Get sample ids
SampleID = data.frame(SampleID = colnames(COUNT),
                 ID = colnames(COUNT)) %>% 
  tidyr::separate(ID, c('A','B','ID'), sep = '_') %>%
  dplyr::select(SampleID, ID) %>%
  dplyr::mutate(ID = as.numeric(ID))

# Get technical metadata
METADATA_TECH_ID <- 'syn6100548'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECH_ID
METADATA_TECH <- read.table(synGet(METADATA_TECH_ID)@filePath,sep=',',header=T, row.names=1) %>%
  group_by(sampleIdentifier) %>%
  top_n(1, -rRNA.rate) %>%
  dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN) %>%
  unique %>%
  ddply(.(barcode), .fun = function(x){
    x = dplyr::mutate(x, batch = paste(batch, collapse = ''), RIN = sum(RIN, na.rm = T))  
  })

# Get clinical metadata
METADATA_CLINICAL_ID <- 'syn6101474'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
METADATA_CLINICAL <- read.csv(synGet(METADATA_CLINICAL_ID)@filePath, header = T)

# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8380379'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)

METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
  gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)
colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))

METADATA <- full_join(METADATA_TECH, METADATA_CLINICAL) %>%
  inner_join(METADATA.PICARD)

GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$ensembl_gene_id
```

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Choose samples with rRNA rate less than 5% and more PF_READS_ALIGNED
METADATA = METADATA %>%
  ddply(.(barcode), .fun = function(x){
    x %>%
      dplyr::filter(PCT_RIBOSOMAL_BASES < 0.05) %>%
      top_n(1, PF_READS_ALIGNED)
  })

# Fix metadata
METADATA = METADATA %>%
  dplyr::mutate(AOD = gsub('\\+','',AOD),
                BrainRegion = factor(BrodmannArea,
                                     levels = c('BM10', 'BM22', 'BM36', 'BM44'),
                                     labels = c('BM10' = 'FP', 'BM22' = 'STG', 
                                                'BM36' = 'PHG', 'BM44' = 'IFG')),
                Diagnosis = 'NONE', 
                RIN2 = RIN^2)
METADATA$Diagnosis[METADATA$CDR <= 1] = 'ND'
METADATA$Diagnosis[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
METADATA$Diagnosis[METADATA$CDR >= 4] = 'SD'
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

METADATA = METADATA %>%
  tidyr::unite(BrainRegion.Diagnosis, BrainRegion, Diagnosis, sep = '.')

METADATA <- METADATA %>%
  filter(!is.na(BrainRegion.Diagnosis), !is.na(PMI), !is.na(RIN))
```

```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$sampleIdentifier, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$sampleIdentifier)

METADATA = as.data.frame(METADATA)
rownames(METADATA) = METADATA$sampleIdentifier
METADATA = METADATA[indToRetain,]

all.counts$MSSM = COUNT
all.mtd$MSSM = METADATA
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn7750270';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)
COUNT[,grep('150_120419', colnames(COUNT))[2]] = NULL

# Get clinical metadata
METADATA.CLINICAL_ID <- 'syn3191087'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.CLINICAL_ID
METADATA.CLINICAL_OBJ <- synGet(METADATA.CLINICAL_ID)
METADATA.CLINICAL <- read.table(METADATA.CLINICAL_OBJ@filePath,sep=',',header=T)

# Get technical covariates
METADATA.TECH_ID <- 'syn4300313'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.TECH_ID
METADATA.TECH_OBJ <- synGet(METADATA.TECH_ID)
METADATA.TECH <- read.table(METADATA.TECH_OBJ@filePath,sep='\t',header=T)

# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8512191'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)

METADATA.PICARD$Sampleid = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
  gsub('Aligned.out/picard.CombinedMetrics.csv','',.)
colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))

# Fix error in technical covariates data
KEY_ID <- 'syn3382527'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = KEY_ID  
KEY <- read.csv(synGet(KEY_ID)@filePath) %>% 
  filter(mrna_data == 1) %>%
  dplyr::select(projid, mrna_id) %>%
  separate(mrna_id, c('a','b','batch'), sep = '_') %>% 
  unite(Sampleid, a, b) %>%
  dplyr::select(-batch) %>%
  unique

# Match technical and clinical covariates
METADATA <- METADATA.TECH %>%
  left_join(METADATA.PICARD) %>%
  dplyr::select(-projid) %>%
  dplyr::left_join(KEY) %>%
  dplyr::left_join(METADATA.CLINICAL)

# Pick higher quality RIN batch for sample 492_120515
METADATA <- METADATA %>%
  group_by(Sampleid) %>%
  top_n(1, RINcontinuous)

GENE.GC = dplyr::select(GENE.PARAM, ensembl_gene_id, percentage_gc_content) %>% unique()
rownames(GENE.GC) = GENE.GC$ensembl_gene_id
```

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  ungroup %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx), !is.na(braaksc), !is.na(ceradsc)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi)) %>%
  filter(!(Sampleid %in% c('380_120503'))) %>%
  as.data.frame()

METADATA$Diagnosis = 'NONE'
METADATA$Diagnosis[METADATA$cogdx == 1 & METADATA$braaksc <= 3 & METADATA$ceradsc > 2] = 'Control'
METADATA$Diagnosis[METADATA$cogdx == 4 & METADATA$braaksc >= 4 & METADATA$ceradsc <= 2] = 'AD'
```

```{r preprocess.data1,cache=TRUE}
# Match covariates to expression data
indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$Sampleid)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$Sampleid
METADATA = METADATA[indToRetain,]

all.counts$ROSMAP = COUNT
all.mtd$ROSMAP = METADATA
```

```{r new.dx}
all.mtd$ROSMAP$nDx = 'OTHER'
all.mtd$ROSMAP$nDx[all.mtd$ROSMAP$cogdx==1 & all.mtd$ROSMAP$braaksc <= 3 & all.mtd$ROSMAP$ceradsc <= 1] = 'CONTROL'
all.mtd$ROSMAP$nDx[all.mtd$ROSMAP$cogdx==4 & all.mtd$ROSMAP$braaksc >= 4 & all.mtd$ROSMAP$ceradsc >= 2] = 'AD'

table(all.mtd$ROSMAP$nDx)

all.mtd$MSSM$nDx = 'OTHER'
all.mtd$MSSM$nDx[all.mtd$MSSM$CDR <= 0.5 & all.mtd$MSSM$bbscore <= 3 & all.mtd$MSSM$NP.1 <= 1] = 'CONTROL'
all.mtd$MSSM$nDx[all.mtd$MSSM$CDR >= 1 & all.mtd$MSSM$bbscore >= 4 & all.mtd$MSSM$NP.1 >= 2] = 'AD'

table(all.mtd$MSSM$nDx, all.mtd$MSSM$BrodmannArea)

table(all.mtd$MAYO$Diagnosis, all.mtd$MAYO$Tissue)
```
